---
title: "Creating custom bootstrap tests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{custom_tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background
p-values can be computed by inverting the corresponding confidence intervals, as described in Section 14.2 of [Thulin (2024)](https://www.modernstatisticswithr.com/mathschap.html#confintequal) and Section 3.12 in [Hall (1992)](https://link.springer.com/book/10.1007/978-1-4612-4384-7). This package contains functions for computing bootstrap p-values in this way. The approach relies on the fact that:

- The p-value of the two-sided test for the parameter theta is the smallest alpha such that theta is not contained in the corresponding 1-alpha confidence interval,
- For a test of the parameter theta with significance level alpha, the set of values of theta that aren't rejected by the two-sided test (when used as the null hypothesis) is a 1-alpha confidence interval for theta.


## Custom hypothesis tests
Bootstrap p-values for hypothesis tests based on `boot` objects can be obtained using the `boot.pval` function.

The example below, based on code from [Section 7.4.2 of Modern Statistics with R](https://www.modernstatisticswithr.com/modchapter.html#a-general-approach), shows how to implement a bootstrap correlation test:

```{r, echo = FALSE}
set.seed(102420)
```


```{r}
library(boot)
library(boot.pval)

# Create a function for computing the correlation given a bootstrap sample:
cor_boot <- function(data, row_numbers, method = "pearson")
{ 
    # Obtain the bootstrap sample:
    sample <- data[row_numbers,]
    
    # Compute and return the statistic for the bootstrap sample:
    return(cor(sample[, 1], sample[, 2], method = method))
}

# We wish to test the correlation between the variables hp and
# drat in the mtcars data.
# Run the resampling:
boot_res <- mtcars |> 
              subset(select = c(hp, drat)) |> 
              na.omit() |> 
              boot(cor_boot, 999)

# Show the bootstrap distribution:
plot(boot_res)

# Compute confidence intervals:
boot.ci(boot_res)

# Compute the bootstrap p-value based on the percentile interval:
boot.pval(boot_res, type = "perc")


# For studentized intervals, we also need to estimate the
# variance of the statistic, e.g. using an inner bootstrap:
cor_boot_student <- function(data, row_numbers, method = "pearson")
{ 
    sample <- data[row_numbers,]
    
    correlation <- cor(sample[, 1], sample[, 2], method = method)
    
    inner_boot <- boot(sample, cor_boot, 100)
    variance <- var(inner_boot$t)

    return(c(correlation, variance))
}

# Run the resampling:
boot_res <- mtcars |> 
              subset(select = c(hp, drat)) |> 
              boot(cor_boot_student, 999)

# Compute the bootstrap p-value based on the studentized interval:
boot.pval(boot_res, type = "stud")
```


The following example is an extensions of that given in the documentation for `boot::boot`:

```{r message = FALSE}
# Hypothesis test for the city data
# H0: ratio = 1.4
ratio <- function(d, w) sum(d$x * w)/sum(d$u * w)
city.boot <- boot(city, ratio, R = 999, stype = "w", sim = "ordinary")
boot.pval(city.boot, theta_null = 1.4)

# Studentized test for the two sample difference of means problem
# using the final two series of the gravity data.
diff.means <- function(d, f)
{
  n <- nrow(d)
  gp1 <- 1:table(as.numeric(d$series))[1]
  m1 <- sum(d[gp1,1] * f[gp1])/sum(f[gp1])
  m2 <- sum(d[-gp1,1] * f[-gp1])/sum(f[-gp1])
  ss1 <- sum(d[gp1,1]^2 * f[gp1]) - (m1 *  m1 * sum(f[gp1]))
  ss2 <- sum(d[-gp1,1]^2 * f[-gp1]) - (m2 *  m2 * sum(f[-gp1]))
  c(m1 - m2, (ss1 + ss2)/(sum(f) - 2))
}
grav1 <- gravity[as.numeric(gravity[,2]) >= 7, ]
grav1.boot <- boot(grav1, diff.means, R = 999, stype = "f",
                   strata = grav1[ ,2])
boot.pval(grav1.boot, type = "stud", theta_null = 0)
```

## References
* Davison, A.C. and Hinkley, D.V. (1997) _Bootstrap Methods and Their Application_. Cambridge University Press.
* Hall P (1992). _The Bootstrap and Edgeworth Expansion_. Springer, New York.
* Thulin, M. (2024) _Modern Statistics with R_. Second edition. Chapman & Hall/CRC Press.

